\section{Lists}
Convenient and predictable list formatting is one of the many advantages of
using LaTeX. Many users of wysiwyg word processors are frequently frustrated by
the software's clumsy attempts to figure out when you intend lists to begin and
end. This is the price of auto-formatting. As a mark-up language, LaTeX gives
you\textit{far} more control over the structure and content of your list. With
a little practice you will find that creating lists in LaTeX is actually a
pleasure when compared to wrestling with your typical "high power" word
processor.
\subsection{List Structures}
 \begin{lstlisting}
itemize
\end{lstlisting}
  \begin{lstlisting}
enumerate
\end{lstlisting}
  \begin{lstlisting}
description
\end{lstlisting}
 Lists often appear in documents, especially academic, as their purpose is
often to present information in a clear and concise fashion. List structures in
LaTeX are simply environments which essentially come in three flavors: ,  and .

All lists follow the basic format:
\begin{lstlisting}
\begin{list_type}
  \item The first item
  \item The second item
  \item The third etc \ldots
\end{list_type}
\end{lstlisting}

All three of these types of lists can have multiple paragraphs per item: just
type the additional paragraphs in the normal way, with a blank line between
each. So long as they are still contained within the enclosing environment,
they will automatically be indented to follow underneath their item.

\subsection{Itemize}
This environment is for your standard bulleted list of items.

\begin{lstlisting}
\begin{itemize}
  \item The first item
  \item The second item
  \item The third etc $\dots$
\end{itemize}
\end{lstlisting} 
\begin{itemize}
\item The first item
\item The second item
\item The third etc $\dots$
\end{itemize}

\subsection{Enumerate}
The enumerate environment is for ordered lists, where by default, each item is
numbered sequentially.

\begin{lstlisting}
\begin{enumerate}
  \item The first item
  \item The second item
  \item The third etc \ldots
\end{enumerate}
\end{lstlisting}
\begin{enumerate}
  \item The first item
  \item The second item
  \item The third etc \ldots
\end{enumerate}

\subsection{Description}
The description environment is slightly different. You can specify the item
label by passing it as an optional argument (although optional, it would look
odd if you didn't include it!). Ideal for a series of definitions, such as a
glossary.

\begin{lstlisting}
\begin{description}
  \item[First] The first item
  \item[Second] The second item
  \item[Third] The third etc \ldots
\end{description}
\end{lstlisting}
\begin{description}
  \item[First] The first item
  \item[Second] The second item
  \item[Third] The third etc \ldots
\end{description}

 \begin{lstlisting}
\\
\end{lstlisting}
  \begin{lstlisting}
\hfill
\end{lstlisting}

 Sometimes you want a description where the text begins on a new line. This
cannot easily be done with . The trick is to use . 
\begin{lstlisting}
\begin{description}
  \item[First] \hfill \\
  The first item
  \item[Second] \hfill \\
  The second item
  \item[Third] \hfill \\
  The third etc \ldots
\end{description}
\end{lstlisting}
\begin{description}
  \item[First] \hfill \\
  The first item
  \item[Second] \hfill \\
  The second item
  \item[Third] \hfill \\
  The third etc \ldots
\end{description}

\subsection{Compacted lists}
\begin{lstlisting}
itemize*
\end{lstlisting}
\begin{lstlisting}
enumerate*
\end{lstlisting}
\begin{lstlisting}
description*
\end{lstlisting}

As you may have noticed, in standard LaTeX document classes, the vertical
spacing between items, and above and below the lists as a whole, is more than
between paragraphs: it may look odd if the descriptions are too short. If you
want tightly-packed lists, use the\texttt{mdwlist} package (included in
the\texttt{mdwtools} bundle), which provides compact, "starred" versions of the
previous environments, i.e. ,  and . They work exactly in the same way, but the
output is more compact.

Alternatively, use the\texttt{memoir} class and with\verb|\tightlists|.

\subsection{Nested Lists}
 \textit{easylist} Latex will happily allow you to insert a list environment
into an existing one (up to a depth of four -- if you need more than four, use
the  package). Simply begin the appropriate environment at the desired point
within the current list. Latex will sort out the layout and any numbering for
you.

\begin{lstlisting}
\begin{enumerate}
  \item The first item
  \begin{enumerate}
    \item Nested item 1
    \item Nested item 2
  \end{enumerate}
  \item The second item
  \item The third etc \ldots
\end{enumerate}
\end{lstlisting}
\begin{enumerate}
  \item The first item
  \begin{enumerate}
    \item Nested item 1
    \item Nested item 2
  \end{enumerate}
  \item The second item
  \item The third etc \ldots
\end{enumerate}

\subsection{Customizing Lists}
Customizing LaTeX is outside the beginners' domain. While not necessarily
difficult in itself, because beginners are already overwhelmed with the array
of commands and environments, moving on to more advanced topics runs the risk
of confusion.

However, since the tutorial is on formatting, I shall still include a brief
guide on customizing lists. Feel free to skip!

\subsubsection{Customizing Line Spacing in Lists}
Inside lists you can redefine some length/dimension variables of latex, for
example using:

\begin{lstlisting}
\begin{itemize}
  \setlength{\itemsep}{1pt}
  \setlength{\parskip}{0pt}
  \setlength{\parsep}{0pt}
  \item first item
  \item second item
\end{itemize}
\end{lstlisting}
\begin{itemize}
  \setlength{\itemsep}{1pt}
  \setlength{\parskip}{0pt}
  \setlength{\parsep}{0pt}
  \item first item
  \item second item
\end{itemize}

Alternatively, to create a unified look in your document you can define your
own enumerate environment:
\begin{lstlisting}
\newenvironment{my_enumerate}
{\begin{enumerate}
  \setlength{\itemsep}{1pt}
  \setlength{\parskip}{0pt}
  \setlength{\parsep}{0pt}}
{\end{enumerate}}
\end{lstlisting}
\newenvironment{my_enumerate}
{\begin{enumerate}
  \setlength{\itemsep}{1pt}
  \setlength{\parskip}{0pt}
  \setlength{\parsep}{0pt}}
{\end{enumerate}}
 
\subsubsection{Customizing Enumerated Lists}
The thing people want to change most often with Enumerated lists are the
counters. Therefore, to go any further, a brief introduction to
LaTeX\textit{counters} is required. Anything that LaTeX automatically numbers,
such as section headers, figures, and itemized lists, there is a counter
associated with it that controls the numbering. 
\begin{lstlisting}
enumi
\end{lstlisting}
  \begin{lstlisting}
enumii
\end{lstlisting}
  \begin{lstlisting}
enumiii
\end{lstlisting}
  \begin{lstlisting}
enumiv
\end{lstlisting}
  \begin{lstlisting}
\setcounter
\end{lstlisting}
  \begin{lstlisting}
\item
\end{lstlisting}
  \begin{lstlisting}
enumi
\end{lstlisting}
There are four individual counters that are associated with itemized lists,
each one represents the four possible levels of nesting, which are called: , ,
, . In order to reset any of these counters in the middle of an enumeration
simply use . The counter is incremented by  before it is printed. For example
to reset  use:

\begin{lstlisting}
\begin{enumerate}
  \setcounter{enumi}{4}
  \item fifth element
\end{enumerate}
\end{lstlisting}

which prints as: 5. fifth element

Each counter also has a default format that dictates how it is displayed
whenever LaTeX needs to print it. Such formats are specified using internal
LaTeX commands:

\begin{tabular}{c c} \hline
 Command
&  Example
\\ \hline
\verb|\arabic|
&  1, 2, 3 ...
\\ \hline
\verb|\alph|
&  a, b, c ...
\\ \hline
\verb|\Alph|
&  A, B, C ...
\\ \hline
\verb|\roman|
&  i, ii, iii ...
\\ \hline
\verb|\Roman|
&  I, II, III ...
\\ \hline
\verb|\fnsymbol|
&  Aimed at footnotes (see below), but prints a sequence of symbols. \\ \hline
\end{tabular}

\begin{lstlisting}
\the
\end{lstlisting}
  \begin{lstlisting}
\theenumi
\end{lstlisting}
 Each counter entity holds various bits of information about itself. To get to
the numbered element, simply use  followed immediately (i.e., no space) by the
name of the counter, e.g., . This is often referred to as
the\textit{representation} of a counter. 

Now, that's most of the technicalities out of the way. To make changes to the
formatting of a given level:
\begin{lstlisting}
\renewcommand{\representation}{\format_command{counter}}
\end{lstlisting}

Admittedly, the generic version is not that clear, so a couple of examples will
clarify:
\begin{lstlisting}
%Redefine the first level
\renewcommand{\theenumi}{\Roman{enumi}}
\renewcommand{\labelenumi}{\theenumi}

%Redefine the second level
\renewcommand{\theenumii}{\Alph{enumii}}
\renewcommand{\labelenumii}{\theenumii}
\end{lstlisting}

The method used above first explicitly changes the format used by the counter.
However, the element that controls the label needs to be updated to reflect the
change, which is what the second line does. Another way to achieve this result
is this:
\begin{lstlisting}
\renewcommand{\labelenumi}{\Roman{enumi}}
\end{lstlisting}

This simply redefines the appearance of the label, which is fine, providing
that you do not intend to cross-reference to a specific item within the list,
in which case the reference will be printed in the previous format. This issue
does not arise in the first example.

Note that you can also add other symbols, such as parentheses and periods,
before and after the counter.  For instance, to create a list indexed by lower
case letters with parentheses before and after the letter, you might enter the
following:
\begin{lstlisting}
\renewcommand{\labelenumi}{(\alph{enumi})}
\end{lstlisting}

\subsubsection{Customizing Itemised Lists}
\begin{lstlisting}
\item[new symbol]
\end{lstlisting}
\begin{lstlisting}
\item[$\star$]
\end{lstlisting}
 Itemized lists are not as complex as they do not need to count. Therefore, to
customize, you simply change the labels. It can be done manually for each entry
with , eg . 

\begin{lstlisting}
\labelitemi
\end{lstlisting}
\begin{lstlisting}
\labelitemii
\end{lstlisting}
\begin{lstlisting}
\labelitemiii
\end{lstlisting}
\begin{lstlisting}
\labelitemiv
\end{lstlisting}
The itemize labels are accessed via , , , , for the four respective levels.
\begin{lstlisting}
\renewcommand{\labelitemi}{\textgreater}
\end{lstlisting}

The above example would set the labels for the first level to a greater than
(>) symbol. Of course, the text symbols available in Latex are not very
exciting. Why not use one of the ZapfDingbat symbols, as described in
the Symbols section.
Or use a mathematical symbol:
\begin{lstlisting}
\renewcommand{\labelitemi}{$\star$}
\end{lstlisting}

Itemized list with tightly set items, that is with no vertical space between
two consecutive items, can be created as follows.
\begin{lstlisting}

\begin{itemize}
  \setlength{\itemsep}{0cm}%
  \setlength{\parskip}{0cm}%
  \item Item opening the list
  \item Item tightly following
\end{itemize}
\end{lstlisting}

\subsubsection{Details of Customizing Lists}
\begin{lstlisting}
\renewcommand
\end{lstlisting}
\begin{lstlisting}
\begin{document}
\end{lstlisting}
Note that it is necessary that the  appears after the  instruction so the
changes made are taken into account. This is needed for both enumerated and
itemized lists.

\subsection{Inline lists}
 \begin{lstlisting}
inparaenum
\end{lstlisting}
Inline lists are a special case as they require the use of the\texttt{paralist}
package which provides the  environment (with an optional formatting
specification in square brackets):

\begin{lstlisting}
...
\usepackage{paralist}

\begin{document}

\textbf{\itshape Inline lists}, which are
sequential in nature, just like enumerated
lists, but are
\begin{inparaenum}[\itshape a\upshape)]
\item formatted within their paragraph;
\item usually labelled with letters; and 
\item usually have the final item prefixed with
`and' or `or',
\end{inparaenum} like this example.
...

\end{lstlisting}

\begin{lstlisting}
\Alph
\end{lstlisting}
\begin{lstlisting}
\alph
\end{lstlisting}
\begin{lstlisting}
\Roman
\end{lstlisting}
\begin{lstlisting}
\roman
\end{lstlisting}
\begin{lstlisting}
\arabic
\end{lstlisting}

To change the styles of the counter, tokens A, a, I, i, and 1 can be used in
the optional argument to produce the counter with one of the styles , , ,  and
. For example:
\begin{lstlisting}
\begin{inparaenum}[(i)]
\end{lstlisting}
 
produces the labels (i), (ii), (iii) ...
