\section{Concepts}
\begin{keywords}
Database, ACID
\end{keywords}
Databases work under a closed world assumption, that means that not all Details
of the actual world are took into consideration. Only attributes relevant to the
task are considered. Another imported aspect is that database give every user
the illusion that he is working alone. Therefor four attributes are defined to
make databases work the way the do. These attributes are called ACID.
\begin{itemize}
\setlength{\itemsep}{1pt}
	\item \textbf{Atomicity}: this means that only one operation at a time will change
the database
	\item \textbf{Consistency}: if the database was in a correct state start before 
any change happend it will be correct state afterwords
	\item \textbf{Isolation}: everyone using the database thinks he is working alone
	\item \textbf{Durability}: once the transaction is done it is save
\end{itemize}

\section{Modelling-Level of Data models}
\begin{itemize}
\setlength{\itemsep}{1pt}
	\item \textbf{conceptual data models} defines the mini-world that is
represented by the database
	\item \textbf{logical data models} defines a specification that is easily
implemented
	\item \textbf{physical model} defines the concrete datastorage
\end{itemize}

\section{Hierarchie Model}
The Hierarchie model defines one dataset and all depending datasets as a
hierarchy. Relation between datasets can have different characteristics.
Think unix-file-system-hierarchy.
\begin{itemize}
\setlength{\itemsep}{1pt}
	\item 1:1 parent-element is bound the child-element
	\item 1:n parent-element is bound to n child-elements
	\item m:n m parent-element are bound to n child-elements
(m $\in [1 \dots n]$)
	\item no cycles are allowed
\end{itemize}

\section{Network Model}
The network-model builds upon the Hierarchie Model. One element can be part of
more than one group.

\section{Flat relational model}
The flat relational model is build from tables hat hab attributes and
column-header. These tables are container for complex datatypes.\\
\begin{tabular}{|c|c|c|c|}
\hline
\multicolumn{4}{|c|}{Participants}\\
\hline
StudentNum & Surname & Name & Semester \\
\hline
4711 & Martha & Mustermann & 3 \\
\hline
42 & Arthur & Dent & 23\\
\hline
\end{tabular}\\
Relations are created through grouping attributes in tables or through
value-based associating.\\
\begin{tabular}{|c|c|}
\hline
\multicolumn{2}{|c|}{Lecture}\\
\hline
LectureNum & Title \\
\hline
48 & Informationsystems \\
\hline
43 & Operation Systems\\
\hline
\end{tabular}\\

\begin{tabular}{|c|c|}
\hline
\multicolumn{2}{|c|}{Selection}\\
\hline
LectureNum & StudentNum \\
\hline
42 & 4711 \\
\hline
43 & 42\\
\hline
\end{tabular}

\section{Nested relational Model}
This allows attribute-values to be relation itself.\\

\scriptsize \begin{tabular}{|c|c|c|c|} 
\hline
GRADE\_REPORT & StudentNum & \multicolumn{2}{|c|}{SectionGrade(SectionID,
Grade)}\\
\hline
& \multirow{2}{*}{17} & 122 & B\\
\cline{3-4}
& & 119 & C\\
\hline
& \multirow{4}{*}{8} & 85 & A\\
& & 92 & A\\
\cline{3-4}
& & 102 & B\\
\cline{3-4}
& & 135 & A\\
\hline
\end{tabular}

\section{Entity-Relationship-Model}
\begin{tikzpicture}
	\node[entity] (emp) {employee};
	\node[attribute] (essn) [above left of=emp,node distance=1.5cm] {\key{Sse}} edge (emp);
	\node[attribute] (bday) [left of=emp,node distance=2.0cm] {BDay} edge (emp);
	\node[attribute] (name) [above of=emp,node distance=1.5cm] {Name} edge (emp);
	\node[attribute] (sName) [above left of=name,node distance=1.1cm] {SName} edge (name);
	\node[attribute] (mName) [above of=name,node distance=1.3cm] {MName} edge (name);
	\node[attribute] (lName) [above right of=name,node distance=1.1cm] {LName} edge (name);
	\node[relationship] (control) [below of=emp,node distance=1.5cm] {Controls} 
		edge node[auto,swap] {1} (emp.-140)
		edge node[auto,swap] {n} (emp.-40);

	\node[entity] (section) [right of=emp,node distance=5.0cm] {section};
	\node[relationship] (works) [right of=emp,node distance=2.5cm] {Works For}
		edge [total] node[auto,swap] {n} (emp)
		edge [total] node[auto,swap] {1} (section);
	\node[relationship] (leads) [below of=works,node distance=1.6cm] {leads}
		edge node[auto,swap] {1} (emp)
		edge [total] node[auto,swap] {1} (section);
	\node[attribute] (name) [above left of=section,node distance=1.5cm] {\key{Name}}
		edge (section);
	\node[attribute] (num) [above of=section,node distance=1.6cm] {\key{Number}}
		edge (section);
	\node[multi attribute] (location) [below of=section,node distance=1.5cm]
		{Location} edge (section);
\end{tikzpicture}

\subsection{Entities}
Entities describe things of the displayed world. Entities can live on there own.
There are two kinds of entities. Strong which have attributes for
identification. Weak entities need extra entities for identification.
\begin{tikzpicture}
	\node[entity] (emp) {Employ};
	\node[ident relationship] (from) [right of=emp, node distance=2.5cm] {belongsTo}
		edge [total] node[auto,swap] {1} (emp);
	\node[weak entity] (fam) [right of=from, node distance=2.5cm] {Family}
		edge [total] node[auto,swap] {n} (from);
\end{tikzpicture}
The employ is a strong entity. It doesn't need any other entity make make itself
identifiable. The family needs an employ entity to be even considered a part of
the model.

\subsection{Attributes}
Attributes describe entities. There are different kinds of attributes.
\begin{itemize}
\setlength{\itemsep}{1pt}
	\item atomar - these attributes can be devided
	\item combined - consist of more than one attribute
	\item univalent - can only appear once
	\item multivalent - can appear more than one time
	\item complex - can consist of different kind of attributes.
	\item key - makes an entity identifiable
	\item partiell key - is part of the key that makes the entity idenfiable
\end{itemize}

\subsection{Relationships}
There are two kinds of relationships \textbf{normal relationships} as well as
\textbf{identifying relationships}.
